---
title: "SVM on IRIS data"
output:
  html_document: default
  pdf_document: 
    keep_tex: yes
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the `iris` data and divide the dataset into a training and a test set (ratio 80/20).

```{r, warning=FALSE}
library(caTools) # needed for sample.split
library(ggplot2) # needed for nicer graphics
library(GGally)  # needed for ggpairs
library(e1071)   # needed for the SVM itself

set.seed(41)        # to make it reproducible

dataset <- iris
str(dataset)
summary(dataset)

split        = sample.split(dataset$Species, SplitRatio = .8)
training_set = subset(dataset, split == TRUE)
test_set     = subset(dataset, split == FALSE)

nrow(training_set)
nrow(test_set)
```

We have 120 data rows on which to train the model, and 30 rows on which to subsequently test it.

Look at the training set more closely, using the plot command `ggpairs`.

```{r}
ggpairs(training_set, ggplot2::aes(colour = Species, alpha = 0.4))
```

This improved pairs-plot contains a large amount of very useful information.

- From the histograms in the bottom 3 rows of the chart of `Petal.length` (3rd column) and `Petal.width` (4th column) we observe a clear separation of the Setosa species.
- However, there is an overlap of the Versicolor and Virginica species. 
- If we look at the scatterplots of `Sepal.Length` vs `Petal.Length` (row 3, column 2) and `Petal.Width` vs `Petal.Length` (row 4, column 3), we can distinctly see a seperator that can be drawn between the groups of species.

Conclusion from the exploratory data analysis:  we can use only `Petal.Width` and `Petal.Length` as parameters for a good model.

We will create two models:

1. containing all 4 parameters,
2. containing only the 2 parameters `Petal.Width` and `Petal.Length`, 

and then compare their performances.

## Scale the features and fit the models

```{r}
training_set[,1:4] = scale(training_set[,1:4])
test_set[,1:4]     = scale(test_set[,1:4])
# model with ALL parameters:
classifier1 = svm(formula = Species~., data = training_set, 
                  type = 'C-classification', kernel = 'radial')
# 2-parameter model:
classifier2 = svm(formula = Species~ Petal.Width + Petal.Length, data = training_set, 
                  type = 'C-classification', kernel = 'radial')
```

## Prediction and accuracy evaluation

We compute the prdiction accuracy directly on the test set.

```{r}
test_pred1 = predict(classifier1, type = 'response', newdata = test_set[-5])
test_pred2 = predict(classifier2, type = 'response', newdata = test_set[-5])

# Compute the confusion matrices
cm1 = table(test_set[,5], test_pred1)
cm2 = table(test_set[,5], test_pred2)
cm1 # Confusion Matrix for all parameters
cm2 # Confusion Matrix for 2-parameter model (Petal Length and Petal Width)
```

## Conclusion

The accuracy of both models is excellent. We conclude that the 2-parameter model suffices.

