---
title: "Clasification with Cross Validation  - use of CARET"
author: 
date: 
output:
  pdf_document:
    keep_tex: yes
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

We show the use of  5 resampling methods on the `iris` data, with the most simple classifier model,  *naive Bayes.* The methods used are:

1. Train-test split.
2. Bootstrap.
3. k-fold CV.
4. k-fold CV with repeats.
5. LOOCV.

## Data

Load the data and split into train-test with a ratio 80-20.

```{r}
# load the libraries
library(caret)
library(klaR) # required for 'naive Bqyes'
# load  `iris` data
data(iris)
# define the train/test split as 80%/20% 
split=0.80
trainIndex <- createDataPartition(iris$Species, p=split, list=FALSE)
data_train <- iris[ trainIndex,]
data_test  <- iris[-trainIndex,]
# fit a "naive bayes" model
model <- NaiveBayes(Species~., data=data_train)
# predictions on the test set
x_test <- data_test[,1:4]
y_test <- data_test[,5]
predictions <- predict(model, x_test)
# print the results
confusionMatrix(predictions$class, y_test)
```

## Resampling by Bootstrap

Sampling with replacement.

```{r}
# define control parameters for the training
train_control <- trainControl(method="boot", number=100)
# fit the model
model <- train(Species~., 
               data=iris, 
               trControl=train_control, 
               method="nb")
# print the results
print(model)
```

## k-fold Cross-Validation

```{r}
# define control parameters for the training
train_control <- trainControl(method="cv", number=10)
# fix tuning parameters of the algorithm
grid <- expand.grid(.fL=c(0), .usekernel=c(FALSE),.adjust=FALSE)
# fit the model
model <- train(Species~., 
               data=iris, 
               trControl=train_control, 
               method="nb", 
               tuneGrid=grid)
# print the results
print(model)
```

## Repeated k-fold Cross-Validation

10-fold with 3 repeats.

```{r}
# define control parameters for the training
train_control <- trainControl(method="repeatedcv", number=10, repeats=3)
# fit the model
model <- train(Species~., 
               data=iris, 
               trControl=train_control, 
               method="nb")
# print the results
print(model)
```

## LOOCV

```{r}
# define control parameters for the training
train_control <- trainControl(method="LOOCV")
# fit the model
model <- train(Species~., 
               data=iris, 
               trControl=train_control, 
               method="nb")
# print the results
print(model)
```

