---
output:
  pdf_document: 
    keep_tex: yes
    highlight: tango
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load librairies and data

Load the necessary libraires and the `iris` data.

```{r, warning=FALSE}
library(tidyverse)  # for ggplot
library(e1071)      # for SVM
set.seed(42)        # to make it reproducible

dataset <- iris
```
## Exploratory data analysis

This has already been performed in Examples \ref{ex:iris-k-nn} and  \ref{ex:iris-svm-feature} above. Please refer back to these.

We just  plot here the measurements $2$ by $2$ (petals followed by sepals).

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point() +
  labs(title = 'Sepal Length vs. Sepal Width')

ggplot(iris, aes(x = Petal.Length, y = Petal.Width, colour = Species)) +
  geom_point() +
  labs(title = 'Petal Length vs. Petal Width')
```

We make the following observations:

1. For sepals, the Setosa specimens are shorter and wider, while the Virginica tends to have longer sepals. There is however a non-trivial overlap among the three.
2. For petals, the Setosas are shorter and narrower than the other two. Virginica tends to have larger petals. There is still some overlap.


## SVM recap.

Datasets presenting an overall separation with some amount of overlap are well adapted to analysis by SVMs. The SVM will provide a classifier for new specimens, trained on the measurements in the given dataset. This is done by computing optimal separator planes (lines in the 2-D case) between two classes that separate all data points into one or the other class. To achieve this, SVM simultaneously deals with three aspects of classification:

1. Class separation, where the algorithm attempts to find an *optimal separating hyperplane* between two or more classes by maximizing the margin between the classes’ closest points. The points lying on the boundaries are called *support vectors*.
2. Overlapping classes, where data points are allowed to be on the wrong side of the discriminating margin, but are weighted to reduce their influence.
3. Nonlinearity, where the data points can in general be separated by nonlinear functions.

The SVM algorithm then finds a solution by formulating the overall task as a quadratic optimization problem that is solved by proven techniques.

## Create the model

Before fitting a model to the Iris dataset, we randomly split it into *training* and *testing* sets, in the ratio 1/3:2/3, using the `sample` command.

```{r}
index <- c(1:nrow(iris))
test.index <- sample(index, size = (length(index)/3))
train_set <- iris[-test.index, ]
test_set  <- iris[ test.index,]
```


## Scale the features and fit the models

It is often advisable to scale the data before fitting the models, especially when the features have very diffrent magnitudes. For the iris data this is not really necessary (though the reader should try, by uncommenting the first 2 lines). Following the analysis of the previous Example \ref{ex:iris-svm-feature}, we examine both the 2-parameter model and the full model containing all $4$ parameters..

We first compute a model with a linear kernel---note that the default for `svm` is the radial kernel.

```{r}
#train_set[,1:4] <- scale(train_set[,1:4])
#test_set[,1:4]  <- scale(test_set[,1:4])

# model with ALL 4 parameters:
classifier1.linear <- svm(formula = Species ~ ., data = train_set, 
                  type = 'C-classification', kernel = 'linear')
classifier1.linear
# 2-parameter model:
classifier2.linear <- svm(formula = Species ~ Petal.Width + Petal.Length, data = train_set, 
                  type = 'C-classification', kernel = 'linear')
classifier2.linear
```

## Prediction and accuracy evaluation

We compute the confusion matrices and evaluate the prediction accuracy on the training data. Note thta this could be considered as a *data crime* (see Section \ref{sec:resample}), but since our aim is to only compare kernels, this is acceptable. Once the best model is selected, we will of course validate it on the *test* data.

```{r}
test_pred1 = predict(classifier1.linear, type = 'response', newdata = test_set[-5])
test_pred2 = predict(classifier2.linear, type = 'response', newdata = test_set[-5])

# Compute the confusion matrices and prediction accuracy
#table(Prediction = predict(svm.model.linear, train),Truth = train$Species)
cm1 = table(test_set[,5], test_pred1)
cm2 = table(test_set[,5], test_pred2)
cm1 # Confusion Matrix for all 4 parameters
cat(sprintf("prediction accuracy for linear kernel = %.2f \n",
            sum(test_set$Species == test_pred1)/nrow(test_set)))
cm2 # Confusion Matrix for 2-parameter model (Petal Length and Petal Width)
cat(sprintf("prediction accuracy for linear kernel = %.2f \n",
            sum(test_set$Species == test_pred2)/nrow(test_set)))
```

We note that the 2-parameter and 4-parameter (full) models have comparable accuracy ($0.94$ vs $0.98$), with a slight advantage to the simpler model. This will however vary if we resample (try this!), so we must perform cross-validation to obtain a stable estimate. This will be done below.

## Tune with other kernels

Now we try the polynomial, radial and sigmoid kernels. These are usually better for more complicated overlaps in multi-class situations.

Let us try the polynomial kernel.

```{r}
classifier1.polynomial <- svm(formula = Species ~ ., data = train_set, 
                  type = 'C-classification', kernel = 'polynomial')
#classifier1.polynomial
classifier2.polynomial <- svm(formula = Species ~ Petal.Width + Petal.Length, data = train_set, 
                  type = 'C-classification', kernel = 'polynomial')
#classifier2.polynomial

test_pred1 = predict(classifier1.polynomial, type = 'response', newdata = test_set[-5])
test_pred2 = predict(classifier2.polynomial, type = 'response', newdata = test_set[-5])

# Compute the confusion matrices and prediction accuracy
cm1 = table(test_set[,5], test_pred1)
cm2 = table(test_set[,5], test_pred2)
cm1 # Confusion Matrix for all parameters
cat(sprintf("prediction accuracy for polynomial kernel = %.2f \n",sum(test_set$Species == test_pred1)/nrow(test_set)))
cm2 # Confusion Matrix for 2-parameter model (Petal Length and Petal Width)
cat(sprintf("prediction accuracy for polynomial kernel = %.2f \n",sum(test_set$Species == test_pred2)/nrow(test_set)))
```

The accuracy is degraded, compared with the linear classifier.

Let us try the radial kernel, which is the default kernel for `svm`.

```{r}
classifier1.radial <- svm(formula = Species ~ ., data = train_set, 
                  type = 'C-classification', kernel = 'radial')
#classifier1.radial
classifier2.radial <- svm(formula = Species ~ Petal.Width + Petal.Length, data = train_set, 
                  type = 'C-classification', kernel = 'radial')
#classifier2.radial

test_pred1 = predict(classifier1.radial, type = 'response', newdata = test_set[-5])
test_pred2 = predict(classifier2.radial, type = 'response', newdata = test_set[-5])

# Compute the confusion matrices and prediction accuracy
cm1 = table(test_set[,5], test_pred1)
cm2 = table(test_set[,5], test_pred2)
cm1 # Confusion Matrix for all parameters
cat(sprintf("prediction accuracy for radial kernel = %.2f \n",sum(test_set$Species == test_pred1)/nrow(test_set)))
cm2 # Confusion Matrix for 2-parameter model (Petal Length and Petal Width)
cat(sprintf("prediction accuracy for radial kernel = %.2f \n",sum(test_set$Species == test_pred2)/nrow(test_set)))
```

Not much difference here, compared to the linear model.

Finally, we try the sigmoid kernel.

```{r}
classifier1.sigmoid <- svm(formula = Species ~ ., data = train_set, 
                  type = 'C-classification', kernel = 'sigmoid')
#classifier1.sigmoid
classifier2.sigmoid <- svm(formula = Species ~ Petal.Width + Petal.Length, data = train_set, 
                  type = 'C-classification', kernel = 'sigmoid')
#classifier2.sigmoid

test_pred1 = predict(classifier1.sigmoid, type = 'response', newdata = test_set[-5])
test_pred2 = predict(classifier2.sigmoid, type = 'response', newdata = test_set[-5])

# Compute the confusion matrices and prediction accuracy
cm1 = table(test_set[,5], test_pred1)
cm2 = table(test_set[,5], test_pred2)
cm1 # Confusion Matrix for all parameters
cat(sprintf("prediction accuracy for sigmoid kernel = %.2f \n",sum(test_set$Species == test_pred1)/nrow(test_set)))
cm2 # Confusion Matrix for 2-parameter model (Petal Length and Petal Width)
cat(sprintf("prediction accuracy for sigmoid kernel = %.2f \n",sum(test_set$Species == test_pred2)/nrow(test_set)))
```

This kernel is slightly less accurate than the linear and radial kernels.

## Intermdiate Conclusion

The accuracy of both models is good, even very good. We conclude that the **2-parameter** model could well suffice, with either a linear or a radial kernel.


## Further optimization and cross-validation

To quote the official `svm`  documentation: ``Parameters of SVM-models usually must be tuned to yield sensible results!''

The `e1071` package of `R`  provides a tuning wrapper function, `tune()`, as well as a function that extracts the best tuned model. The `tune` function performs a grid search over the supplied parameter ranges. The performance measure is the classification error (or mean-squared error for regression). Ten-fold cross-validation is performed by default, though this can be modified. For additional diagnostics, the cost function's evolution can be plotted as a function of the parameters to obtain a performance profile. This will be done in the following example.

Warning: the CPU time for the tuning cycle can be quite long, depending on the dimensions of the parameter range vectors.

```{r}
tuned1.svm <- tune.svm(Species~., data = train_set, kernel = 'linear',
         gamma = 0.25, cost = 2^seq(-6, 4, 2))
#        gamma = seq(1/2^nrow(iris),1, .01), cost = 2^seq(-6, 4, 2))
tuned1.svm
tuned2.svm <- tune.svm(Species~Petal.Width + Petal.Length, data = train_set, 
                       kernel = 'linear',
         gamma = 2^(-1:1), cost = 2^seq(-6, 4, 2))
#        gamma = seq(1/2^nrow(iris),1, .01), cost = 2^seq(-6, 4, 2))
tuned2.svm
```

Using these optimized values we create a new model and recompute the confusion matrix and classification accuracy.

```{r}
tuned1.linear <- svm(formula = Species ~ ., data = train_set, 
                  type = 'C-classification', kernel = 'linear',gamma = 0.25, cost = 0.25)
tuned2.linear <- svm(formula = Species ~ Petal.Width + Petal.Length, data = train_set, 
                  type = 'C-classification', kernel = 'linear',gamma = 0.25, cost = 1.0)

test_pred1 = predict(tuned1.linear, test_set)
  #predict(tuned1.linear, type = 'response', newdata = test_set[-5])
test_pred2 = predict(tuned2.linear, type = 'response', newdata = test_set[-5])

# Compute the confusion matrices and prediction accuracy
cm1 = table(test_set[,5], test_pred1)
cm2 = table(test_set[,5], test_pred2)
cm1 # Confusion Matrix for all parameters
cat(sprintf("prediction accuracy for best 4-parameter model = %.2f \n",sum(test_set$Species == test_pred1)/nrow(test_set)))
cm2 # Confusion Matrix for 2-parameter model (Petal Length and Petal Width)
cat(sprintf("prediction accuracy for best 2-parameter model = %.2f \n ",sum(test_set$Species == test_pred2)/nrow(test_set)))
```

Using the `best.svm()`  method, which is is really just a wrapper for `tune.svm(...)$best.model`.)

```{r}
best.svm1 <- best.svm(Species~. , data = train_set, kernel = 'linear',
                      gamma = seq(0,1, .05), cost = 2^seq(-6, 4, 2))
best.svm1
table(Prediction = predict(best.svm1, train_set), Truth = train_set$Species)
best.svm.pred1 <- predict(best.svm1, test_set)
table(Prediction = best.svm.pred1, Truth = test_set$Species)
cat(sprintf("prediction accuracy for best 4-parameter model = %.2f \n",sum(test_set$Species == best.svm.pred1)/nrow(test_set)))
```

Excellent precision for the full model.

```{r}
best.svm2 <- best.svm(Species ~ Petal.Width + Petal.Length, data = train_set, kernel = 'linear')
best.svm2
table(Prediction = predict(best.svm2, train_set), Truth = train_set$Species)
best.svm.pred2 <- predict(best.svm2, test_set)
table(Prediction = best.svm.pred2, Truth = test_set$Species)
cat(sprintf("prediction accuracy for best 2-parameter model = %.2f \n",sum(test_set$Species == best.svm.pred2)/nrow(test_set)))
```

The 2-parameter model performs admirably, though with a slightly inferior precision.

